Calculating impedances for CMA and CA areas from the GSS surveys

### 2024-11-08 data-raw/01-preprocessing-source-files.Rmd
Update documentation

---
title: "Reading source files and preparing data sets"
author: "Bruno Santos, Mahdis Moghadasi & Antonio Paez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

This Rmarkdown file is part of the
[**ActiveCA**](https://github.com/dias-bruno/ActiveCA) package. This
package is one of the products developed by the [*Mobilizing
Justice*](https://mobilizingjustice.ca/)[^1].

[^1]: The Mobilizing Justice project is a multidisciplinary and
    multi-sector collaboration with the objective of understand and
    address transportation poverty in Canada and to improve the
    well-being of Canadians at risk of transport poverty. The Social
    Sciences and Humanities Research Council (SSRHC) has provided
    funding for the project, which was created by an unprecedented
    alliance of academics from various Canadian provinces and
    institutions, transportation firms, and nonprofit organizations

The main objective of the *ActiveCA* package is to identify the most
appropriate impedance functions for active transportation modes for
various destinations and time periods in Canada. To do achieve our
objective, we analyzed historical data from the [General Social Survey
(GSS)](https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=5221)
from 1986 to 2015 in Canada to calculate the impedance function for
cycling and walking trips.

@@

@@

Additionally, our analysis will focus on developing impedance functions considering only the population who lives in larger urban population centre (a Census Metropolitan Area (CMA) or Census Agglomeration (CA))^[For definitions of CMA and CA check the this [website](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/az/Definition-eng.cfm?ID=geo009).]. We decided to apply this restriction because the travel behavior among residents of the CMA and CA areas, in terms of active travel, tends to be very different from those outside these large urban population centers. The variable that informs if the respondent resides in a CMA or CA region is the variable named `LUC_RST`, present in the Main File survey. According to the Statistics Canada, `LUC_RST` = 1 indicates that the resident is from a CMA or CA region.

@@

```{r echo = FALSE}
walking_2015 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
cycling_2015 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
walking_2010 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
cycling_2010 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
walking_2005 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
cycling_2005 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
walking_1998 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
cycling_1998 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
walking_1992 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
cycling_1992 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r echo = FALSE}
walking_1986 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

>>>>

---
title: "Reading source files and preparing data sets"
author: "Bruno Santos, Mahdis Moghadasi & Antonio Paez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: activeca.bib
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

This Rmarkdown file is part of the
[**ActiveCA**](https://github.com/dias-bruno/ActiveCA) package. The main objective of the *ActiveCA* package is to identify the most
appropriate impedance functions for active transportation modes for
various destinations and time periods in Canada. To do achieve this objective, we analyzed historical data from the [General Social Survey
(GSS)](https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=5221)
from 1986 to 2015 in Canada to calculate the impedance function for
cycling and walking trips.

@@

## Suggested reading

-   Soukhov, A., Páez, A. (2024). Accessibility analysis for planning
    applications (Report No. MJ-A2-0002). Mobilizing Justice.
    <https://github.com/soukhova/MJ-Accessibility-Blogs>

@@

@@

```{r include=FALSE}
walking_2015 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
cycling_2015 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
walking_2010 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
cycling_2010 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
walking_2005 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
cycling_2005 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
walking_1998 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
cycling_1998 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
walking_1992 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
cycling_1992 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

@@

```{r include=FALSE}
walking_1986 |>
  filter(orig_label == "Home" & dest_label == "Home")
```

### 2024-11-08 data-raw/02-calculating-impedances.Rmd
Update documentation

---
title: "Calculating impedances for walking in cycling"
author: "Bruno Santos, Mahdis Moghadasi & Antonio Paez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
editor_options: 
  markdown: 
    wrap: 72
---

@@

This Rmarkdown file is part of the
[**ActiveCA**](https://github.com/dias-bruno/ActiveCA) package. This
package is one of the products developed by the [*Mobilizing
Justice*](https://mobilizingjustice.ca/)[^1].

The main objective of the *ActiveCA* package is to identify the most
appropriate impedance functions for active transportation modes for
various destinations and time periods in Canada. To do achieve our
objective, we analyzed historical data from the [General Social Survey
(GSS)](https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=5221)
from 1986 to 2015 in Canada to calculate the impedance function for
cycling and walking trips.

@@

To exemplify the process of selecting the best impedance function, we will use the example of the 1986 GSS dataset, already pre-processed in the `00-Reading-Source-Files.Rmd`: 
```{r}
load(file = paste0(here(), "/data/walking_1986.Rda"))
```

@@

Analyzing the histogram above, we can identify that the just a few values of duration are higher than 1 hundred minutes.  To better visualize the variable distribution, let's use an other type of graph, the boxplot graph:

@@

```{r}
length(walking_1986$DURATION)
sum(walking_1986$DURATION > 100)
```
Visualizing the new summary:
```{r}
walking_1986 <- walking_1986 %>% filter(DURATION <= 100)
summary(walking_1986$DURATION)
```

@@

The next chunks of code of this Rmarkdown explains the process of calibrating PDF
functions for the Walking dataset of the 1986 GSS.

@@

Creating a data frame with the all empirical travel times:

@@

@@

Obtaining the theoretical impedance function for the each destination mode:
```{r}
for(destino in unique(tld_empirical$destination)){
  assign(paste0("f_1986_Walking_dest", destino), 
         test_distributions_weighted(
           x = tld_empirical[tld_empirical$destination == destino, ]$DURATION,
           weights = tld_empirical[tld_empirical$destination == destino, ]$WGHT_EPI))
}
```

@@

Actually, the `Log-normal distribution` was the best fuction for the others two destination from the survey: `2` (`Work or school`) and `3` (`Other's home`).

@@

Building the theoretical values for each destination based on the best
impedance function for each destination:

@@

@@

```{r travel-time-plots}
par(mfrow = c(2, round(length(unique(tld_theoretical$destination))/2)))  # Arrange plots in 2 columns

# Loop through each destination
for (destination in unique(tld_empirical$destination)) {
  # Subset empirical and theoretical data for the current destination
  empirical_subset <- tld_empirical[tld_empirical$destination == destination, ]
  theoretical_subset <- tld_theoretical[tld_theoretical$destination == destination, ]

  # Create histogram for empirical data
  hist(empirical_subset$DURATION, freq = FALSE, col = "grey70", breaks = 20, main = paste(empirical_subset$dest_label[1]),
       xlab = "Travel time from home to work (min)", ylab = "Density")

  # Add theoretical curve
  lines(theoretical_subset$t, theoretical_subset$f)
}

@@

@@

```{r}
gss_episodes <- gss_episodes %>%
  filter(DURATION <= 100 & DURATION > 0)

gss_episodes <- gss_episodes %>%
  group_by(MODE, YEAR, destination) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  filter(count > 1)

@@

```{r}

@@

          assign(paste0("f_",year, "_", mode, "_dest", destino),
               test_distributions_weighted(
               x = gss_subset$DURATION,
               weights = round(gss_subset$WGHT_EPI)))
        }
    }
  }
}
```

@@

```

```{r}
gss_impedances <- gss_episodes[,c("PUMFID","DURATION","LOCATION",
                  "origin","destination","orig_label",
                  "dest_label","YEAR","MODE","f_name",
                  "f","est_1","est_2","loglik","aic","bic")]
```

>>>>

---
title: "Calculating impedances for CMA and CA areas from the GSS surveys"
author: "Bruno Santos, Mahdis Moghadasi & Antonio Paez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: activeca.bib
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

@@

This Rmarkdown file is part of the
[**ActiveCA**](https://github.com/dias-bruno/ActiveCA) package.The main objective of the *ActiveCA* package is to identify the most
appropriate impedance functions for active transportation modes for
various destinations and time periods in Canada. To do achieve our
objective, we analyzed historical data from the [General Social Survey
(GSS)](https://www23.statcan.gc.ca/imdb/p2SV.pl?Function=getSurvey&SDDS=5221)
from 1986 to 2015 in Canada to calculate the impedance function for
cycling and walking trips.

@@

Soukhov and Páez [-@MJ-A2-0002] explained the process of selecting the best impedance function in detail in a report released by the [*Mobilizing Justice* partnership](https://mobilizingjustice.ca/)[^1] using data from the city of Hamilton, Ontario. Subsequently, Santos and Páez (2024) adapted the methodology to create [**CommuteCA**](https://github.com/dias-bruno/CommuteCA), a package to obtain accessibility measures for any Canadian city using the 2021 Population Census as the main data source. Here, we will apply the methodology developed by Santos and Páez (2024) to obtain the impedance functions of the GSS surveys.

To exemplify the process of selecting the best impedance function, we will use the example of the 1986 GSS dataset, already pre-processed in the `00-Reading-Source-Files.Rmd`:

```{r}
load(file = paste0(here(), "/data/walking_1986.Rda"))
```

@@

Analyzing the histogram above, we can identify that the just a few values of duration are higher than 1 hundred minutes.  To better visualize the variable distribution, let's use an other type of graph, the box plot graph:

@@

```{r}
length(walking_1986$DURATION)
sum(walking_1986$DURATION > 100)
```
Visualizing the new summary:
```{r}
walking_1986 <- walking_1986 %>% filter(DURATION <= 100)
summary(walking_1986$DURATION)
```
After excluding trip longer than 100 minutes duration, the minimum duration value for walking trips is `1` minute (same as before),  while the maximum is `95` minutes. The new average duration for walking trips is `17.03` minutes and the median is `10` minutes.

@@

@@

The next chunks of code of this Rmarkdown explains the process of calibrating PDF
functions for the Walking dataset of the 1986 GSS. Creating a data frame with the all empirical travel times:

@@

The code below introduces two functions: `test_distributions_weighted` and `lowest_aic`. The first takes the time values as input and applies them to the distributions mentioned above, obtaining an AIC value for each distribution. In turn, `lowest_aic` compares the AIC values of each distribution and selects the function with the minimum AIC (the distribution with the best fit considering the time values provided).

@@

The next chunk selects the best impedance function, using the `test_distributions_weighted` and `lowest_aic` functions for each destination and transportation mode (in this case, walking since the 1986 survey does not have cycling as an mode option). The name pattern is as follows: `f_year_mode_dest` + `destination`, which means that `f_1986_Walking_dest1` displays the best function for the year 1986, walking mode, destination 1 (Home).

```{r}
for(destino in unique(tld_empirical$destination)){
  assign(paste0("f_1986_Walking_dest", destino), 
         test_distributions_weighted(
           x = tld_empirical[tld_empirical$destination == destino, ]$DURATION,
           weights = tld_empirical[tld_empirical$destination == destino, ]$WGHT_EPI))
}
```

@@

Actually, the `Log-normal distribution` was the best function for the others two destination from the survey: `2` (`Work or school`) and `3` (`Other's home`).

@@

Obtaining the theoretical values for each destination based on the best
impedance function for each destination:

@@

Visualizing the theoretical values (first 10 rows):
```{r}
tld_theoretical[1:10,]
```

@@

```{r travel-time-plots}
par(mfrow = c(2, round(length(unique(tld_theoretical$destination))/2)))

# Loop through each destination
for (destination in unique(tld_empirical$destination)) {
  empirical_subset <- tld_empirical[tld_empirical$destination == destination, ]
  theoretical_subset <- tld_theoretical[tld_theoretical$destination == destination, ]

hist(empirical_subset$DURATION, freq = FALSE, col = "grey70", breaks = 20, main = paste(empirical_subset$dest_label[1]),
       xlab = "Travel time from home to work (min)", ylab = "Density")

  lines(theoretical_subset$t, theoretical_subset$f)
}

@@

After exemplifying the process of selecting the best impedance function for the 1986 GSS survey, we will obtain the function for all GSS cycles. First, we'll load the GSS episodes pre-processed in the first rmarkdown:

@@

Additionally, our analysis will focus on developing impedance functions considering only the population who lives in larger urban population centre (a Census Metropolitan Area (CMA) or Census Agglomeration (CA))^[For definitions of CMA and CA check the this [website](https://www12.statcan.gc.ca/census-recensement/2021/ref/dict/az/Definition-eng.cfm?ID=geo009).]. We decided to apply this restriction because the travel behavior among residents of the CMA and CA areas, in terms of active travel, tends to be very different from those outside these large urban population centers. The variable that informs if the respondent resides in a CMA or CA region is the variable named `Pop_centre`, obtained after some processing made with he Main File survey.

```{r}
gss_episodes <- gss_episodes %>%
  filter(DURATION <= 100 & 
           DURATION > 0 & 
           YEAR > 1986 & 
           Pop_centre %in% c("CMA/CA", "CMA"))
```

Furthermore, to calculate the function, we need a combination of `mode`, `year` and `destination` greater than 1, which means that we need at least two episodes to be able to fit a function. Consequently, combinations with a greater number of episodes will exhibit better functions in terms of AIC and will better represent the empirical values.

```{r}
gss_episodes <- gss_episodes %>%
  group_by(MODE, YEAR, destination) %>%
  mutate(count = n()) %>%
  ungroup() %>%
  filter(count > 1)

@@

Similar as we did before, the next chunk selects the best impedance function, using the `test_distributions_weighted` and `lowest_aic` functions for each destination, transportation mode and year. The name pattern is as follows: `f_year_mode_dest` + `destination`, which means that `f_1986_Walking_dest1` displays the best function for the year 1986, walking mode, destination 1 (Home).
/000
```{r}

@@

          assign(paste0("f_",year, "_", mode, "_dest", destino),
               test_distributions_weighted(
               x = gss_subset$DURATION,
               weights = round(gss_subset$WGHT_EPI)))
        }
    }
  }
}
```
After selecting the best function for every combination of destination, transportation mode and year, the next two chunks will apply the functions in all surveys, adding the information with the best function and its parameters, AIC/BIC/loglik values, and count of episodes that were used to calculate the function:

@@

```
Creation of a new dataset called `gss_impedances`, to keep the original episode files as they are after the pre-processing stage:

```{r}
gss_impedances <- gss_episodes[,c("PUMFID","DURATION","LOCATION",
                  "origin","destination","Province","Pop_centre","orig_label",
                  "dest_label","YEAR","MODE","f_name",
                  "f","est_1","est_2","loglik","aic","bic","count")]
```
